The Galactic Delay-Time Distribution of Type Ia Supernovae:\\ A Chemical Evolution Perspective

### 2023-08-21 src/tex/ms.tex
[showyourwork] overleaf sync

\usepackage{showyourwork}
\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{layouts}
\usepackage{xcolor}
% \usepackage{siunitx}

@@

\begin{deluxetable*}{lll}
    \tablecaption{Sample selection parameters from APOGEE DR17\label{tab:sample-selection}}
    \tablehead{
        \colhead{Parameter} & \colhead{Range} & \colhead{Notes}
    }
    \startdata
        $\log g$            & $1.0 < \log g < 3.8$          & Select giants only \\
        $T_{\rm eff}$       & $3500 < T_{\rm eff} < 5500$ K & Reliable temperature range \\
        $S/N$               & $S/N > 80$                    & Required for accurate stellar parameters \\
        ASPCAPFLAG Bits     & $\notin$ 23                   & Remove stars flagged as bad \\
        EXTRATARG Bits      & $\notin$ 0, 1, 2, 3, or 4     & Select main red star sample only \\
    \enddata
\end{deluxetable*}

@@

\begin{deluxetable}{Ccl}
    \tablecaption{A summary of parameters and their fiducial values for our one-zone chemical evolution models.\label{tab:onezone-parameters}}
    \tablehead{
        \colhead{Parameter} & \colhead{Value} & \colhead{Description}
    }
    \startdata
    \Delta t        & 10 Myr    & Time-step size \\
    t_{\rm max}     & 13.2 Gyr  & Maximum simulation time \\
    \eta            & 2.15      & Outflow mass-loading factor \\
    \tau_*          & 2.0 Gyr   & Star formation efficiency timescale \\
    \dot M_r        & continuous    & Recycling rate \\
    t_D             & 40 Myr    & Minimum SN Ia delay time \\
    M_{\rm gas,0}   & 0         & Initial gas mass \\
    \dot M_*(t)     & Equation \ref{eq:insideout-sfh}  & Star formation rate \\
    \tau_{\rm rise} & 2 Gyr     & SFR rise timescale \\
    \tau_{\rm sfh}  & 15.1 Gyr  & SFR exponential timescale
    \enddata
\end{deluxetable}

\subsection{The SN Ia Delay-Time Distribution}
\label{sec:onezone-dtd}

@@

\begin{deluxetable}{lll}
    \tablecaption{Median Parameter Uncertainties\label{tab:uncertainties}}
    \tablehead{
        \colhead{Parameter} & \colhead{Median Uncertainty} & \colhead{Source}
    }
    \startdata
        [Fe/H] & $9.2\times10^{-3}$ & APOGEE DR17 \\
        $\rm [O/Fe]$ & $1.8\times10^{-2}$ & APOGEE DR17 \\
        log(Age/Gyr) & $0.12$ & \citet{Leung2023-Ages}
        % Age/Gyr & 29\% & astroNN
    \enddata
\end{deluxetable}

@@

To quantify the similarity between the MDFs generated by VICE and the observed APOGEE MDF, we compute the Kullback-Leibler (KL) divergence \citep{KullbackLeibler1951}, defined as
\begin{equation}
\label{eq:kl-divergence}
D_{\rm{KL}}(P \parallel Q) = \int_{-\infty}^{\infty} p(x) \log\Big(\frac{p(x)}{q(x)}\Big) dx
\end{equation}
for distributions $P$ and $Q$ with probability density functions $p(x)$ and $q(x)$. In this case, $P$ is the APOGEE MDF, $Q$ is the model MDF, and $x$ is [Fe/H]. For each model SFH and DTD, we compute $D_{\rm{KL}}$ in 18 different galactocentric zones defined by radial bins $3-5$ kpc, $5-7$ kpc, $7-9$ kpc, $9-11$ kpc, $11-13$ kpc, and $13-15$ kpc and absolute $z$-height bins $0-0.5$ kpc, $0.5-1$ kpc, and $1-2$ kpc. The KL divergence for the entire model is taken to be the average of $D_{\rm{KL}}$ for each zone, weighted by the number of APOGEE stars in each zone.

@@

Picking the team.

% TODO make table with numerical scores (KL divergences, etc) for supplementary material
% TODO explain that precise statistical understanding of these scores is outside the scope of the paper

@@

Figure \ref{fig:onezone-greggio05} shows the results of one-zone chemical evolution models with the \citet{Greggio2005-AnalyticalRates} DTDs. We assume $\eta=2.5$, $\tau_*=2$ Gyr, an inside-out SFH evaluated at a radius of 8 kpc, a continuous recycling approximation, and a minimum SN Ia delay of 40 Myr. We compare the SD, DD WIDE, and DD CLOSE schemes to standard power-law, broken power-law, and exponential DTDs. The SD and DD CLOSE DTDs follow nearly identical tracks in [O/Fe] vs [Fe/H]; however, their distributions on [O/Fe] differ at the low end. The SD DTD follows an exponential with a 1.5 Gyr timescale, whereas the DD CLOSE DTD is well-approximated by a broken power-law with an initial plateau of 350 Myr and a subsequent declining slope of -1.1. The WIDE prescription is likewise best approximated by a broken power-law, but with a longer plateau width of 1 Gyr. In all cases, the difference between the analytical DTD and the nearest broken power-law or exponential is likely too small to be observationally detectable, so in our multi-zone models we implement the more generic functions in lieu of the analytical DTDs.

@@

We fit a sech$^2$ function \citep{Spitzer1942} to the distribution of midplane distances $z_{\rm final}$. Vertical migration away from the midplane does not affect the chemical evolution simulation, but we do use $z_{\rm final}$ in our analysis. The probability density function (PDF) of $z_{\rm final}$ given some scale height $h_z$ is
\begin{equation}
    {\rm PDF}(z_{\rm final}) = \frac{1}{4 h_z} {\rm sech}^2\Big(\frac{z_{\rm final}}{2 h_z}\Big)
    \label{eq:sech-pdf}
\end{equation}
and the corresponding cumulative distribution function (CDF) is
\begin{equation}
    {\rm CDF}(z_{\rm final}) = \frac{1}{1 + e^{-z_{\rm final} / h_z}}.
    \label{eq:sech-cdf}
\end{equation}
We fit the above function to the distributions of $z_{\rm final}$ in h277 in varying bins of $\tau$ and $R_{\rm final}$ and found that $h_z$ is best described by the function
\begin{equation}
    h_z = (0.25\,{\rm kpc}) 
    e^{\frac{\tau-5\,{\rm Gyr}}{7.0\,{\rm Gyr}}}
    e^{\frac{R_{\rm final}-8\,{\rm kpc}}{6.0\,{\rm kpc}}}.
    \label{eq:scale-height}
\end{equation}

@@

\begin{figure*}
    \centering
    \includegraphics{figures/radial_migration.pdf}
    \caption{The distribution of final radius $R_{\rm final}$ as a function of formation radius $R_{\rm form}$ and age for the h277 analogue (top row) and Gaussian sampling scheme (bottom row). From left to right, star particles are binned by formation annulus, from the inner disk (far left column) to the Solar annulus (center column) to the outer disk (far right column). Within each panel, colored curves represent the different age bins, ranging from the youngest stars (dark blue) to the oldest (dark red). All distributions are normalized so that the area under the curve is 1, and have been smoothed by a boxcar function of width 0.5 kpc. The vertical dotted black lines indicate the bounds of each bin in $R_{\rm form}$; stars within that region of the distribution have not migrated outside their birth annulus over their lifetime.}
    \label{fig:radial-migration}
    \script{radial_migration.py}
\end{figure*}

\begin{figure*}
    \centering
    \includegraphics{figures/midplane_distance.pdf}
    \caption{Similar to Figure \ref{fig:radial-migration} but for the distribution of final midplane distance $z_{\rm final}$ as a function of final radius and age. From left to right, star particles are binned by \textit{final} annulus. All distributions have been smoothed by a boxcar function of width 0.1 kpc.}
    \label{fig:midplane-distance}
    \script{midplane_distance.py}
\end{figure*}

>>>>

\usepackage{showyourwork}
\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{layouts}
\usepackage{xcolor}
\usepackage{upgreek}
% \usepackage{siunitx}

@@

\begin{deluxetable*}{lll}
    \tablecaption{Sample selection parameters from APOGEE DR17\label{tab:sample-selection}}
    \tablehead{
        \colhead{Parameter} & \colhead{Range} & \colhead{Notes}
    }
    \startdata
        $\log g$            & $1.0 < \log g < 3.8$          & Select giants only \\
        $T_{\rm eff}$       & $3500 < T_{\rm eff} < 5500$ K & Reliable temperature range \\
        $S/N$               & $S/N > 80$                    & Required for accurate stellar parameters \\
        ASPCAPFLAG Bits     & $\notin$ 23                   & Remove stars flagged as bad \\
        EXTRATARG Bits      & $\notin$ 0, 1, 2, 3, or 4     & Select main red star sample only \\
        Age                 & $\sigma_{\rm Age} < 40\%$     & Age uncertainty from \citet{Leung2023-Ages}
    \enddata
\end{deluxetable*}

@@

\begin{deluxetable}{Ccl}
    \tablecaption{A summary of parameters and their fiducial values for our one-zone chemical evolution models.\label{tab:onezone-parameters}}
    \tablehead{
        \colhead{Quantity} & \colhead{Value} & \colhead{Description}
    }
    \startdata
    \Delta t        & 10 Myr    & Time-step size \\
    t_{\rm max}     & 13.2 Gyr  & Maximum simulation time \\
    \eta            & 2.15      & Outflow mass-loading factor \\
    \tau_*          & 2.0 Gyr   & Star formation efficiency timescale \\
    \dot M_r        & continuous    & Recycling rate \\
    t_D             & 40 Myr    & Minimum SN Ia delay time \\
    M_{\rm gas,0}   & 0         & Initial gas mass \\
    \dot M_*(t)     & Equation \ref{eq:insideout-sfh}  & Star formation rate \\
    \tau_{\rm rise} & 2 Gyr     & SFR rise timescale \\
    \tau_{\rm sfh}  & 15.1 Gyr  & SFR exponential timescale
    \enddata
\end{deluxetable}

@@

\begin{deluxetable*}{Ccl}
    \tablecaption{A summary of parameters and their fiducial values for our multi-zone chemical evolution models. We omit parameters which have been left unchanged from \citet{Johnson2021-Migration}; see their Table 1 for details.\label{tab:multizone-parameters}}
    \tablehead{
        \colhead{Quantity} & \colhead{Value} & \colhead{Description}
    }
    \startdata
        t_D                 & 40 Myr                                & Minimum SN Ia delay time \\
        \Delta R_{\rm gal}  & Equation \ref{eq:radial-migration}    & Change in radius due to stellar migration (see Appendix \ref{app:migration}) \\
        z                   & Equation \ref{eq:sech-pdf}            & Distance from Galactic midplane at present day (see Appendix \ref{app:migration}) \\
        f_{\rm EB}(t|R_{\rm gal})   & Equation \ref{eq:earlyburst-ifr}  & Time-dependence of the infall rate in the early-burst model \\
        f_{\rm TI}(t|R_{\rm gal})   & Equation \ref{eq:twoinfall-ifr}   & Time-dependence of the infall rate in the two-infall model \\
        \uptau_{*,\rm EB}(t|R_{\rm gal})  & Equation \ref{eq:earlyburst-conroy22} & Time-dependence of the SFE timescale in the early-burst model \\
    \enddata
\end{deluxetable*}

\begin{deluxetable}{lll}
    \tablecaption{Median Parameter Uncertainties\label{tab:uncertainties}}
    \tablehead{
        \colhead{Parameter} & \colhead{Median Uncertainty} & \colhead{Source}
    }
    \startdata
        [Fe/H] & $9.2\times10^{-3}$ & APOGEE DR17 \\
        $\rm [O/Fe]$ & $1.8\times10^{-2}$ & APOGEE DR17 \\
        log(Age/Gyr) & $0.10$ & \citet{Leung2023-Ages}
        % Age/Gyr & 29\% & astroNN
    \enddata
\end{deluxetable}

@@

@@

We've run a lot of models and need some semi-quantitative way to say that some models do better than others. We compare our multi-zone outputs to APOGEE across five diagnostics: the [Fe/H] DF, [O/Fe] DF, [Fe/H]--[O/Fe] plane, age--[O/Fe] plane, and bimodality in the [O/Fe] DF. We perform statistical tests between APOGEE and simulation outputs in each region of the galaxy, then average the scores together (weighted by the number of APOGEE targets in each galactic region) to obtain a single numerical score for each simulation. Note that we are not trying to use these scores to fit our models to the data, nor are we trying to rule out any SFH in general. We use these scores to get a handle on which combinations of SFH and DTD are favorable or disfavorable in certain regimes.

To quantify the difference between the abundance distributions generated by VICE and those observed in APOGEE, we compute the Kullback-Leibler (KL) divergence \citep{KullbackLeibler1951}, defined as
\begin{equation}
    D_{\rm{KL}}(P \parallel Q) \equiv \int_{-\infty}^{\infty} p(x) \log\Big(\frac{p(x)}{q(x)}\Big) dx
    \label{eq:kl-divergence}
\end{equation}
for distributions $P$ and $Q$ with probability density functions $p(x)$ and $q(x)$. A KL divergence of 0 indicates that the two distributions contain equal amounts of information. In this case, $P$ is the APOGEE DF, $Q$ is the model DF, and $x$ is either [Fe/H] or [O/Fe]. For each model SFH and DTD, we compute $D_{\rm{KL}}$ in 18 different Galactocentric regions: bins in $R_{\rm gal}$ of $3-5$ kpc, $5-7$ kpc, $7-9$ kpc, $9-11$ kpc, $11-13$ kpc, and $13-15$ kpc, and bins in $|z|$ of $0-0.5$ kpc, $0.5-1$ kpc, and $1-2$ kpc. The KL divergence for the entire model is taken to be the average of $D_{\rm{KL}}$ for each region weighted by the number of APOGEE stars in that region.

\citet{PerezCruz2008-KLTest2D} present a method for estimating the KL divergence between two continuous, multivariate samples using a $k$-nearest neighbor estimate. We implement this method to calculate the divergence between the multi-zone models and APOGEE data in [Fe/H]---[O/Fe] space.\footnote{We got the implementation from \url{https://mail.python.org/pipermail/scipy-user/2011-May/029521.html}.}

We employ a different test in the age---[O/Fe] plane. [motivated by the large age uncertainties?] As shown in Figures \ref{fig:age-ofe-sfh} and \ref{fig:age-ofe-dtd}, in each Galactic region we bin the model outputs and data into bins of [O/Fe] with a width of 0.05 dex. We define the RMS median age difference for the region as
\begin{equation}
    \Delta\tau_{\rm RMS} \equiv \sqrt{\frac{\sum_k \Delta\tau_k^2 n_{{\rm L23},k}}{n_{\rm L23,tot}}}
    \label{eq:age-ofe-score}
\end{equation}
where $\Delta\tau_k=\rm{med}(\tau_{\rm VICE})-\rm{med}(\tau_{\rm L23})$ is the difference between the mass-weighted median star particle age in VICE and the median stellar age from \citet{Leung2023-Ages} in bin $k$, $n_{{\rm L23},k}$ is the number of stars from the \citet{Leung2023-Ages} age sample in bin $k$, and $n_{\rm L23,tot}$ is the total number of stars in the sample in that Galactic region. As before, the score for the model as a whole is the average of $\Delta \tau_{\rm RMS}$ across all regions, weighted by the number of stars in the age sample in each region.

The results of these tests are presented in Table \ref{tab:results}. Instead of reporting the numerical scores for each simulation output, in the column corresponding to a particular diagnostic we simply write \yes, \meh, or \no, which corresponds to a score in the top, middle, or bottom third out of all of our simulations, respectively. A \yes indicates that particular combination of model SFH and DTD performed better than most when compared to the APOGEE sample for that diagnostic.

@@

Figure \ref{fig:onezone-greggio05} shows the results of one-zone chemical evolution models with the \citet{Greggio2005-AnalyticalRates} DTDs. We assume $\eta=2.5$, $\tau_*=2$ Gyr, an inside-out SFH evaluated at a radius of 8 kpc, a continuous recycling approximation, and a minimum SN Ia delay of 40 Myr. We compare the SD, DD WIDE, and DD CLOSE schemes to standard power-law, broken power-law, and exponential DTDs. The SD and DD CLOSE DTDs follow nearly identical tracks in [O/Fe] vs [Fe/H]; however, their distributions on [O/Fe] differ at the low end. The SD DTD follows an exponential with a 1.5 Gyr timescale, whereas the DD CLOSE DTD is well-approximated by a broken power-law with an initial plateau of 300 Myr and a subsequent declining slope of -1.1. The WIDE prescription is likewise best approximated by a broken power-law, but with a longer plateau width of 1 Gyr. In all cases, the difference between the analytical DTD and the nearest broken power-law or exponential is likely too small to be observationally detectable, so in our multi-zone models we implement the more generic functions in lieu of the analytical DTDs.

@@

We fit a sech$^2$ function \citep{Spitzer1942} to the distribution of midplane distances $z$. Vertical migration away from the midplane does not affect the chemical evolution simulation, but we do use $z$ in our analysis. The probability density function (PDF) of $z$ given some scale height $h_z$ is
\begin{equation}
    {\rm PDF}(z) = \frac{1}{4 h_z} {\rm sech}^2\Big(\frac{z}{2 h_z}\Big)
    \label{eq:sech-pdf}
\end{equation}
and the corresponding cumulative distribution function (CDF) is
\begin{equation}
    {\rm CDF}(z) = \frac{1}{1 + e^{-z / h_z}}.
    \label{eq:sech-cdf}
\end{equation}
We fit the above function to the distributions of $z$ in h277 in varying bins of $\tau$ and $R_{\rm final}$ and found that $h_z$ is best described by the function
\begin{equation}
    h_z = (0.25\,{\rm kpc}) 
    e^{\frac{\tau-5\,{\rm Gyr}}{7.0\,{\rm Gyr}}}
    e^{\frac{R_{\rm final}-8\,{\rm kpc}}{6.0\,{\rm kpc}}}.
    \label{eq:scale-height}
\end{equation}

@@

\begin{figure*}
    \centering
    \includegraphics{figures/radial_migration.pdf}
    \caption{The distribution of final radius $R_{\rm final}$ as a function of formation radius $R_{\rm form}$ and age for the h277 analogue (top row) and Gaussian sampling scheme (bottom row). From left to right, star particles are binned by formation annulus, from the inner disk (far left column) to the Solar annulus (center column) to the outer disk (far right column). Within each panel, colored curves represent the different age bins, ranging from the youngest stars (dark blue) to the oldest (dark red). In the top row, we exclude bins with fewer than 100 unique analogue IDs for clarity. All distributions are normalized so that the area under the curve is 1, and have been smoothed by a boxcar function of width 0.5 kpc. The vertical dotted black lines indicate the bounds of each bin in $R_{\rm form}$; stars within that region of the distribution have not migrated outside their birth annulus over their lifetime.}
    \label{fig:radial-migration}
    \script{radial_migration.py}
\end{figure*}

\begin{figure*}
    \centering
    \includegraphics{figures/midplane_distance.pdf}
    \caption{Similar to Figure \ref{fig:radial-migration} but for the distribution of final midplane distance $z_{\rm final}$ as a function of final radius and age. From left to right, star particles are binned by \textit{final} annulus. In the top row, we exclude bins with fewer than 500 unique analogue IDs for clarity. All distributions have been smoothed by a boxcar function of width 0.1 kpc.}
    \label{fig:midplane-distance}
    \script{midplane_distance.py}
\end{figure*}
